\pkg{cubble}: An R Package for Structuring Spatio-temporal Data

### 2022-02-04 main.tex
make notes and small changes till section 4.2

\Abstract{
The abstract of the article.
}

\Keywords{spatio-temporal data, \proglang{R}}
\Plainkeywords{spatio-temporal data, R}

@@

\usepackage{amsmath} \usepackage{array}

@@

The rest of the paper will be divided as follows: {[}complete when the
paper structure is more solid{]}

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:illu-input} shows four examples of this. No matter which form
the data is in, these formats share some common components that
characterise spatio-temporal data. A spatial identifier (\texttt{id} in
the diagram) is the unique identifier of each site. The temporal
identifier (\texttt{t} in the diagram) prescribes the time stamp each
site is recorded. Coordinates, comprising of latitude and longitude
(\texttt{lon} and \texttt{lat} in the diagram), locates each site on the
map. These identifiers will be the building blocks for the data
structure introduced below. Other variables in the data can be
categorised into two groups: spatial variables that are invariant at
each time stamp for every site, i.e.~the name or code of the weather
station and temporal variables that varies with time.

@@

{\centering \includegraphics[width=1\linewidth,height=0.15\textheight]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by site
and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\texttt{rowwise\_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\texttt{map} syntax when working with list-column.

@@

To use functionalities from cubble, data analysts first need to create a
cubble. \texttt{as\_cubble} create a \pkg{cubble} by supplying the three
key components: \texttt{key} as the spatial identifier; \texttt{index}
as the temporal identifier; and a vector of \texttt{coords} in the order
of longitude first and then latitude. The naming of \texttt{key} and
\texttt{index} follows the convention in the \pkg{tsibble} package. The
cubble created by default is in the nested form. Below is an example of
creating a cubble: \newline

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

There are a few information in the \pkg{cubble} header: the name of the
\texttt{key} variable, bbox, and also the name of variable nested in the
\texttt{ts} column. In this example, each site is identifier is
\texttt{id} and the number in the bracket means there are 5 unique
\texttt{id} in this dataset. The bbox in the second row gives the range
of the coordinates. The temporal variables are all nested in the
\texttt{ts} column, but it could be useful to know the name these
variables. The third row in the cubble header shows these names and in
this example this includes: precipitation, \texttt{prcp}, maximum
temperature, \texttt{tmax}, and minimum temperature, \texttt{tmin}.

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

Notice here that the third line in the header now shows the name of
spatial variables rather than the temporal variables.

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, analysts will need to go back and forth between the
nested and long cubble. The \texttt{stretch()} function introduced in
the previous section switches a nested cubble into a long cubble and
function \texttt{tamp()} is its inverse function to switch a long cubble
back to the nested cubble:

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

As a final data output for modelling or visualisation, spatio-temporal
data is usually expected to be in a single table. Function
\texttt{migrate()} moves the spatial variables from the \texttt{spatial}
attribute into the long cubble:

@@

In this workflow described above, data objects come into cubble in the
nested form, then various operations on the spatial and temporal
dimension can go back and forth between the nested and long form, and
finally, the data will come out of cubble in the long form for further
modelling or visualisation.

@@

\pkg{cubble} is also compatible with \pkg{tsibble} in the sense that the
original list-column can be a \texttt{tbl\_ts} object. Duplicates and
gaps should be first checked before structuring the data into a cubble.
If the input data is a \pkg{tsibble} object, the long form cubble is
also a \pkg{tsibble} where users can directly apply time series
operations.

@@

Spatial locations can have further grouping structure either in nature
(i.e.~countries are nested in continents), or calculated by some
clustering algorithms. Rather than investigate the read from individual
locations, group character summarised from multiple locations can give a
clearer picture of the local dynamic. In cubble, \texttt{switch\_key()}
can be used to create a new level of grouping of spatial location by
specifying a clustering variable. Figure \ref{fig:illu-hier} illustrates
the relationship of cubbles at site and cluster level, in both the long
and nested form. By specifying
\texttt{cluster\_nested\ \textless{}-\ station\_nested\ \%\textgreater{}\%\ switch\_key(key\ =\ cluster)},
the cubble re-defines the cubble key from \texttt{id} in
\texttt{station\_nested} to \texttt{cluster} in
\texttt{cluster\_nested}. All the spatial variables variant to
\texttt{cluster} are now nested into a \texttt{.val} column and cluster
level variables can be computed in the same fashion as station level
variables in \texttt{station\_nested}. Following the same principal of
\texttt{stretch}, the long form cluster level cubble expands the time
series variables with both indices: \texttt{cluster} and \texttt{id}.
All the cluster level variables are stored separately for recovering the
nested from from the long form.

@@

{\centering \includegraphics[width=1\linewidth]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.003}

@@

The cubble structure fits in naturally with the interactive graphic
pipeline discussed in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works with
the two forms in a cubble, where a time series plot is created with the
long cubble and a map is created with the nested cubble. When a user
action is captured from the map, the site will be activated in the
nested cubble. Then, the nested cubble will communicate to the long
cubble to activate all the observations with the same \texttt{id}. The
long cubble will then highlight the activated series in the time series
plot.

The linking is also available from the time series plot to the map. The
selection on the time series is through selecting the point on the time
series and once a point is selected, it will be activated in the long
cubble. All the observations that share the same \texttt{id}, either in
the long and nested cubble, are then activated. This includes other
points in the same time series in the long cubble and the corresponding
observation of site in the nested cubble. These activated observations
will then being reflected in the updated plots and Diagram
\ref{fig:illu-interactive-2} in the Appendix illustrates this process.

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map. In \proglang{R}, \texttt{GGally} implements the glyph map
through the \texttt{glyphs()} function, which outputs a data frame with
calculated position (\texttt{gx}, \texttt{gy}, \texttt{gid}) of each
point on the time series given the major and minor xy variable using
linear algebra (Equation 1 and 2 in \citet{Wickham2012-yr}). The data
can then be piped into \texttt{ggplot} to create the glyph map:

@@

While this calculation can be seen as part of the \texttt{setup\_data()}
in a \texttt{ggproto}. A re-implementation of the glyph map as
\texttt{GeomGlyph} has been made in \texttt{cubble} to create the glyph
map with \texttt{geom\_glyph}:

@@

Polar glyph map can be specify as a parameter, \texttt{polar\ =\ TRUE},
in the \texttt{geom\_glyph()}, along with \texttt{width} and
\texttt{height} in either absolute or relative value. Global and local
scale can also be controlled by the parameter \texttt{global\_rescale},
which default to \texttt{TRUE}. Reference box and line can be added with
separate \texttt{geom\_glyph\_box()} and \texttt{geom\_glyph\_line()}.

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world and the dataset
\texttt{weatherdata::historical\_tmax} extracts the historical maximum
temperature recorded for 236 Australian stations. The data
\texttt{historical\_tmax} is already presented as a cubble, with
\texttt{id} as the key, \texttt{date} as the index, and
\texttt{c("longitude",\ "latitude")} as the coordinates. Other variables
include \texttt{elevation}, \texttt{name}, \texttt{wmo\_id},
\texttt{first\_year}, and \texttt{last\_year} in the nested form and
\texttt{tmax} in the long form. This example compares the maximum
temperature in two periods: 1971 - 1975 and 2016 - 2020 for stations in
Victoria and New South Wales.

Stations in the two states can be subsetted on the station number:
Australia GHCN station number starts with \texttt{ASN00} and followed by
the \href{http://www.bom.gov.au/climate/cdo/about/site-num.shtml}{Bureau
of Meteorology (BOM) station number}, where the 2nd and 3rd digit (7th
and 8th in the GHCN number) denote the state a station belongs to. New
South Wales stations start from 46 to 75 and Victoria stations then
follow from 76 to 90. Extracting Victoria and New South Wales stations
is a filter operation in the spatial dimension and hence is operated in
the nested form:

@@

The five year window is chosen to remove the effect of a particular year
and the historical period of 1971 - 1975 is used since all the stations
have records from 1970. This following chunk filters on records in the
two study periods and summarises the maximum temperature into monthly
measure. \texttt{stretch()} is used to convert the nested form cubble
into the long form for these operations:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>% 
+   stretch() %>%
+   filter(lubridate::year(date) %in% c(1971:1975, 2016:2020)) %>%
+   mutate(month = lubridate::month(date), 
+          group = as.factor(ifelse(lubridate::year(date) > 2015, 
+                                   "2016 ~ 2020", "1971 ~ 1975"))) %>%
+   group_by(month, group) %>%
+   summarise(tmax = mean(tmax, na.rm = TRUE))
\end{CodeInput}
\end{CodeChunk}

A data quality issue with GHCN data is that while the first and last
year of each series is provided, years missing in this period is not
reported. There are a few stations which don't have records during 1971
- 1975 and these stations are filtered out by examining whether the
summarised \texttt{tmax} has a total of 24 months. This is again a
station-wise operation and is operated in the nested form, which is
switched to from the long form by \texttt{tamp()}:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>% 
+   tamp() %>%
+   filter(nrow(ts) == 24) 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>%   
+   stretch() %>%
+   migrate(latitude, longitude)
\end{CodeInput}
\end{CodeChunk}

Figure \ref{fig:basic-manip} shows the glyph map made with the data
after the wangling above. One issue with this map is that the similar
pattern shown from a few nearby stations around Sydney (151E, 34S) and
New castle (152E, 33S) can be distracting. Aggregation is a useful
technique to observe the general pattern of a collection of series and
will be the topic of the next example.

@@

\caption[Glyph map of the mean maximum temperature by month for Victoria and New South Wales weather stations]{Glyph map of the mean maximum temperature by month for Victoria and New South Wales weather stations. On the top left corner is a detailed legend for station Cobar highlighted in the black box. Compared to 1971 - 1975,  the period 2016 - 2020 sees an increase of mean maximum temperature in spring and summer in Australia (end and beginning of the year). A larger increase during the first three months of the year is observed at stations with a lower latitude.}\label{fig:basic-manip}
\end{figure}
\end{CodeChunk}

@@

There are 639 stations that record daily precipitation in Australia and
these series will overlay on top of each other in a glyph map.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map. This example shows how to
organise data at both level with the cubble function
\texttt{switch\_key()}.

A simple kmean algorithm based on the distance matrix is used here to
create 20 clusters. This creates \texttt{station\_nested} as a station
level nested cubble with a cluster column indicating which group each
station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

\begin{CodeChunk}
\begin{CodeInput}
R> station_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [639]: nested form
# bbox:     [113.53, -43.66, 153.64, -10.05]
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name             wmo_id ts               cluster
  <chr>       <dbl> <dbl> <dbl> <chr>             <dbl> <list>             <int>
1 ASN00001006 -15.5  128.   3.8 wyndham aero      95214 <tibble [1,827 ~       3
2 ASN00001007 -13.8  126.   6   troughton island  94102 <tibble [1,827 ~       3
3 ASN00001018 -16.4  126. 546   mount elizabeth   94211 <tibble [1,460 ~      16
4 ASN00001019 -14.3  127.  23   kalumburu         94100 <tibble [1,827 ~       3
5 ASN00001020 -14.1  126.  51   truscott          95101 <tibble [1,827 ~       3
# ... with 634 more rows
\end{CodeOutput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], prcp [dbl]
  id          name                lat  long ts         type   dist group n_match
  <chr>       <chr>             <dbl> <dbl> <list>     <chr> <dbl> <int>   <int>
1 405234      SEVEN CREEKS @ D~ -36.9  146. <tibble [~ river  6.15     5      21
2 ASN00082042 strathbogie       -36.8  146. <tibble [~ clim~  6.15     5      21
3 404207      HOLLAND CREEK @ ~ -36.6  146. <tibble [~ river  8.54    10      21
4 ASN00082170 benalla airport   -36.6  146. <tibble [~ clim~  8.54    10      21
5 230200      MARIBYRNONG RIVE~ -37.7  145. <tibble [~ river  6.17     6      19
6 ASN00086038 essendon airport  -37.7  145. <tibble [~ clim~  6.17     6      19
7 406213      CAMPASPE RIVER @~ -37.0  145. <tibble [~ river  1.84     1      18
8 ASN00088051 redesdale         -37.0  145. <tibble [~ clim~  1.84     1      18
\end{CodeOutput}
\end{CodeChunk}

>>>>

\Abstract{
Sptio-temporal data naturally has two levels of observations: site and
time level. {[}problem{]}. In this paper, a new data structure,
\texttt{cubble}, is proposed for manipulating and visualising
spatio-temporal data.
}

\Keywords{spatio-temporal data, \proglang{R}, climate data}
\Plainkeywords{spatio-temporal data, R, climate data}

@@

\usepackage{amsmath} \usepackage{array} \usepackage{xcolor}

@@

The rest of the paper will be divided as follows: Section 2 presents the
workflow of data manipulation in cubble. Section 3 explains how cubble
deals with some advanced considerations including data with hierarchical
structure, data matching, and how cubble fits with existing static and
interactive visualisation tools. Section 4 gives examples of the
features introduced in the previous two sections with climate and
hydrology data. An example on how cubble handle Netcdf data is also
provided. Section 5 concludes the paper

@@

Spatio-temporal data usually come in various forms and Figure
\ref{fig:illu-input} shows four examples of this. No matter which form
the data is in, these formats share some common components that
characterise the spatio-temporal data. A spatial identifier (\texttt{id}
in the diagram) is the unique identifier of each site. The temporal
identifier (\texttt{t} in the diagram) prescribes the time stamp each
site is recorded. Coordinates, comprising of latitude and longitude
(\texttt{lon} and \texttt{lat} in the diagram), locates each site on the
map. These identifiers will be the building blocks for the data
structure introduced below. Other variables in the data can be
categorised into two groups: spatial variables that are invariant at
each time stamp for every site, i.e.~the name or code of the weather
station, and temporal variables that varies with time.

@@

{\centering \includegraphics[width=1\linewidth,height=0.4\textheight]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.001}

@@

In a cubble, there are two forms: nested form and long form, and Figure
\ref{fig:illu-cubble} sketches the two forms along with the associated
attributes. The decision on which form to use is output-oriented,
meaning analysts need to first think about whether the output of a
particular operation is identified only by the spatial identifier, or a
combination of spatial and temporal identifier. The nested cubble is
suitable for working with operations that are only identified by the
site and this type of operation can be a pure manipulation of spatial
variables, or a summary of temporal variables by site (i.e.~the output
of counting the number of raining day is only identified by sites and
hence should be performed with the nested form). Underneath the nested
form, a cubble is built from a row-wise dataframe (\texttt{rowwise\_df})
where each site forms a separate group. This structure simplifies the
calculation that involves temporal variables by avoiding the use of
\texttt{map} syntax when working with list-column.

@@

\textcolor{red}{To use functionalities from cubble, data analysts first need to create a cubble.}
\texttt{as\_cubble()} creates a \pkg{cubble} by supplying the three key
components: \texttt{key} as the spatial identifier; \texttt{index} as
the temporal identifier; and a vector of \texttt{coords} in the order of
longitude first and then latitude. The naming of \texttt{key} and
\texttt{index} follows the convention in the \pkg{tsibble} package. The
cubble created by default is in the nested form. Below is an example of
creating a cubble: \newline

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_nested <- cubble::climate_flat %>%
+   as_cubble(key = id, index = date, coords = c("long", "lat"))
R> cubble_nested
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

There are a few information in the \pkg{cubble} header: the name of the
\texttt{key} variable (\texttt{id}) and its number (5), bounding box
(\texttt{{[}115.97,\ -32.94,\ 133.55,\ -12.42{]}}), and also the name of
variable nested in the \texttt{ts} column with its type
(\texttt{date\ {[}date{]},\ prcp\ {[}dbl{]},\ tmax\ {[}dbl{]},\ tmin\ {[}dbl{]}}).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_long <- cubble_nested %>% stretch(ts)
R> cubble_long
\end{CodeInput}
\begin{CodeOutput}
# cubble:  date, id [5]: long form
# bbox:    [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# spatial: lat [dbl], long [dbl], elev [dbl], name [chr], wmo_id [dbl]
  id          date        prcp  tmax  tmin
  <chr>       <date>     <dbl> <dbl> <dbl>
1 ASN00009021 2020-01-01     0  31.9  15.3
2 ASN00009021 2020-01-02     0  24.9  16.4
3 ASN00009021 2020-01-03     6  23.2  13  
4 ASN00009021 2020-01-04     0  28.4  12.4
5 ASN00009021 2020-01-05     0  35.3  11.6
# ... with 1,825 more rows
\end{CodeOutput}
\end{CodeChunk}

Notice here that the third line in the header now shows the name and
type of spatial variables
(\texttt{lat\ {[}dbl{]},\ long\ {[}dbl{]},\ elev\ {[}dbl{]},\ name\ {[}chr{]},\ wmo\_id\ {[}dbl{]}}).

@@

Manipulation on the spatial and temporal dimension can be an iterative
process. Many times, analysts will need to go back and forth between the
nested and long cubble. \texttt{tamp()} inverses the \texttt{stretch()}
that introduced in the previous section, to switch from a long cubble to
the nested cubble:

\begin{CodeChunk}
\begin{CodeInput}
R> cubble_back <- cubble_long %>% tamp()
R> cubble_back
\end{CodeInput}
\begin{CodeOutput}
# cubble:   id [5]: nested form
# bbox:     [115.97, -32.94, 133.55, -12.42]- check gap on long and lat
# temporal: date [date], prcp [dbl], tmax [dbl], tmin [dbl]
  id            lat  long  elev name           wmo_id ts                
  <chr>       <dbl> <dbl> <dbl> <chr>           <dbl> <list>            
1 ASN00009021 -31.9  116.  15.4 perth airport   94610 <tibble [366 x 4]>
2 ASN00010311 -31.9  117. 179   york            94623 <tibble [366 x 4]>
3 ASN00010614 -32.9  117. 338   narrogin        94627 <tibble [366 x 4]>
4 ASN00014015 -12.4  131.  30.4 darwin airport  94120 <tibble [366 x 4]>
5 ASN00015131 -17.6  134. 220   elliott         94236 <tibble [366 x 4]>
\end{CodeOutput}
\end{CodeChunk}

@@

\textcolor{red}{As a final data output for modelling or visualisation, spatio-temporal data is usually expected to be in a single table.}
Function \texttt{migrate()} moves the spatial variables from the
\texttt{spatial} attribute into the long cubble:

@@

@@

\pkg{cubble} is also compatible with \pkg{tsibble} where each element
list-column in the nested form can be a \texttt{tbl\_ts} object.
\textcolor{red}{Duplicates and gaps should be first checked before structuring the data into a cubble.}
If the input data is a \pkg{tsibble} object, the long form cubble is
also a \pkg{tsibble} where users can directly apply time series
operations. \textcolor{red}{more experiment with tsibble}

@@

Spatial locations can have grouping structure that is either inherent to
the data (i.e.~countries are nested in continents), or formed by
clustering. Rather than analysing variables in the site level,
summarised variables in the cluster level can give a clearer picture of
\textcolor{red}{local dynamic}. In cubble, \texttt{switch\_key()} can be
used to create a new level of grouping of spatial location by specifying
a clustering variable. Figure \ref{fig:illu-hier} illustrates the
relationship of cubbles at site and cluster level, in both the long and
nested form. \textcolor{red}{unify the use of site/ station}. By
specifying
\texttt{cluster\_nested\ \textless{}-\ station\_nested\ \%\textgreater{}\%\ switch\_key(key\ =\ cluster)},
the cubble re-defines the cubble key from \texttt{id} in
\texttt{station\_nested} to \texttt{cluster} in
\texttt{cluster\_nested}. All the spatial variables variant to
\texttt{cluster} are now nested into a \texttt{.val} column and cluster
level variables can be computed in the same fashion as station level
variables in \texttt{station\_nested}. Following the same principal of
\texttt{stretch}, the long form cluster level cubble expands the time
series variables with both indices: \texttt{cluster} and \texttt{id}.
All the cluster level variables are stored separately for recovering the
nested from from the long form.

@@

{\centering \includegraphics[width=1\linewidth,height=0.4\textheight]{/Users/sherryzhang/Documents/research/paper-cubble/figures/diagram-keynotes/diagram-keynotes.003}

@@

Cubble fits in naturally with the interactive graphic pipeline discussed
in the literature
\citep{buja1988elements, buja1996interactive, sutherland2000orca, xie2014reactive, cheng2016enabling}.
Diagram \ref{fig:illu-interactive} illustrates how linking works from
the map to the time series with the two forms of a cubble. The map and
time series plot is associated with the nested or long cubble
respectively and when a user action is captured on the map, the site
will be activated in the nested cubble (left). The nested cubble will
communicate to the long cubble to activate all the observations with the
same \texttt{id} (middle). The long cubble will then highlight the
activated series in the time series plot (right).

The linking is also available from the time series plot to the map. The
selection(s) on the time series is through selecting the point(s) on the
time series and once a point is selected, it will be activated in the
long cubble. All the observations that share the same \texttt{id} are
then activated and this includes other points in the same time series in
the long cubble and the corresponding observation of site in the nested
cubble. These activated observations will then being reflected in the
updated plots and Diagram \ref{fig:illu-interactive-2} in the Appendix
illustrates this process.

@@

Glyph map \citep{Wickham2012-yr} plots the time series as single glyph
on the map and in \proglang{R}, \texttt{GGally} implements the glyph map
through the \texttt{glyphs()} function. The function construct a data
frame with calculated position (\texttt{gx}, \texttt{gy}, \texttt{gid})
of each point on the time series using linear algebra (Equation 1 and 2
in \citet{Wickham2012-yr}). The data can then be piped into
\texttt{ggplot} to create the glyph map as:

@@

A re-implementation of the glyph map as a ggproto \texttt{GeomGlyph} has
been made in the \texttt{cubble} package so that the glyph map can be
created with \texttt{geom\_glyph()}:

@@

Polar glyph map can be specify as a parameter, \texttt{polar\ =\ TRUE},
in the \texttt{geom\_glyph()}, along with \texttt{width} and
\texttt{height} in either absolute or relative value. Global and local
scale can be controlled by the parameter \texttt{global\_rescale}, which
default to \texttt{TRUE} for global scaling. Reference box and line can
be added with separate \texttt{geom\_glyph\_box()} and
\texttt{geom\_glyph\_line()}.

@@

Global Historical Climatology Network (GHCN) provides daily climate
measures from stations across the world and the dataset
\texttt{weatherdata::historical\_tmax} extracts the maximum temperature
for 236 Australian stations. All the stations start recording from year
1969 with the longest back to year 1859. The data
\texttt{historical\_tmax} is already presented as a cubble, with
\texttt{id} as the key, \texttt{date} as the index, and
\texttt{c("longitude",\ "latitude")} as the coordinates. This example
compares the maximum temperature in two periods: 1971 - 1975 and 2016 -
2020 for stations in Victoria and New South Wales.

Stations in the two states can be subsetted on the station number:
Australia GHCN station number starts with ``ASN00'' and followed by the
\href{http://www.bom.gov.au/climate/cdo/about/site-num.shtml}{Bureau of
Meteorology (BOM) station number}, where the 2nd and 3rd digit (7th and
8th in the GHCN number) define the state of the station. New South Wales
stations start from 46 to 75 and Victoria stations follow from 76 to 90.
Filtering Victoria and New South Wales stations is an operation in the
spatial dimension and hence uses the nested form:

@@

Filtering for the period 1971: 1975 and 2016:2020 is an operation on the
time dimension and the nested cubble needs to be switched to the long
cubble by \texttt{stretch()}:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>% 
+   stretch() %>%
+   filter(lubridate::year(date) %in% c(1971:1975, 2016:2020)) 
\end{CodeInput}
\end{CodeChunk}

A monthly average is used for both periods to smooth the maximum
temperature and it is again an operation on the time dimension:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>%
+   mutate(month = lubridate::month(date), 
+          group = as.factor(ifelse(lubridate::year(date) > 2015, 
+                                   "2016 ~ 2020", "1971 ~ 1975"))) %>%
+   group_by(month, group) %>%
+   summarise(tmax = mean(tmax, na.rm = TRUE))
\end{CodeInput}
\end{CodeChunk}

There are a few stations which don't have records during 1971 - 1975 and
further investigation shows that while the first and last year of each
series is recorded, the missing years in this period is not reported.
These stations are filtered out by examining whether the summarised time
series has a total of 24 months. The long cubble needs to be switched to
the nested form for this spatial operation using \texttt{tamp()}:

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>% tamp() %>% filter(nrow(ts) == 24) 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeInput}
R> tmax <- tmax %>% stretch() %>% migrate(latitude, longitude)
\end{CodeInput}
\end{CodeChunk}

\texttt{tmax} can then be supplied to \texttt{geom\_glyph()} for the
glyph map in Figure \ref{fig:basic-manip} with a station inset on the
top left corner:

\begin{CodeChunk}
\begin{CodeInput}
R> nsw_vic <- ozmaps::abs_ste %>% 
+   filter(NAME %in% c("Victoria", "New South Wales"))
R> 
R> ggplot() + 
+   geom_sf(data = nsw_vic, 
+           fill = "transparent", color = "grey", linetype = "dotted") + 
+   geom_glyph(data = tmax, 
+              aes(x_major = longitude, x_minor = month, 
+                  y_major = latitude, y_minor = tmax,
+                  group = interaction(id, group), color = group),
+              width = 1, height = 0.5) +
+   ...
\end{CodeInput}
\end{CodeChunk}

@@

\caption[A glyph map of the mean maximum temperature by month for Victoria and New South Wales weather stations in Australia]{A glyph map of the mean maximum temperature by month for Victoria and New South Wales weather stations in Australia. On the top left corner is an insetted plot of station Cobar, which is highlighted in the black box.}\label{fig:basic-manip}
\end{figure}
\end{CodeChunk}

@@

In the previous example, there has already been some overlapping of the
glyphs for a few stations near (151E, 34S) and (152E, 33S) and this will
be a problem when mapping more stations in the national level.
Aggregation can be helpful to group series into clusters before
visualising the cluster with glyph map. This example shows how to
organise data at both level with \texttt{switch\_key()}. The dataset
used is \texttt{weatherdata::climate\_full}, which records daily
precipitation and maximum/ minimum temperature for 639 stations in
Australia from 2016 to 2020.

A simple kmean algorithm based on the distance matrix is used here to
create 20 clusters. This creates \texttt{station\_nested} as a station
level nested cubble with a cluster column indicating the group each
station belongs to. More complex algorithms can be used for other
problem, as long as there is a mapping from each station to a cluster.

\begin{CodeChunk}
\begin{CodeInput}
R> station_nested <- weatherdata::climate_full %>% 
+   mutate(cluster = ...)
\end{CodeInput}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{CodeOutput}
# cubble:   id [8]: nested form
# bbox:     [144.52, -37.73, 146.06, -36.55]
# temporal: date [date], matched_var [dbl]
  id          name                lat  long type    dist group ts        n_match
  <chr>       <chr>             <dbl> <dbl> <chr>  <dbl> <int> <list>      <int>
1 405234      SEVEN CREEKS @ D~ -36.9  146. river   6.15     5 <tibble ~      21
2 404207      HOLLAND CREEK @ ~ -36.6  146. river   8.54    10 <tibble ~      21
3 ASN00082042 strathbogie       -36.8  146. clima~  6.15     5 <tibble ~      21
4 ASN00082170 benalla airport   -36.6  146. clima~  8.54    10 <tibble ~      21
5 230200      MARIBYRNONG RIVE~ -37.7  145. river   6.17     6 <tibble ~      19
6 ASN00086038 essendon airport  -37.7  145. clima~  6.17     6 <tibble ~      19
7 406213      CAMPASPE RIVER @~ -37.0  145. river   1.84     1 <tibble ~      18
8 ASN00088051 redesdale         -37.0  145. clima~  1.84     1 <tibble ~      18
\end{CodeOutput}
\end{CodeChunk}
